---
title: "Assignment Notebook"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: false
---

```{r, message = FALSE, warning=FALSE}
require(kableExtra)
require(gt)
require(gtsummary)
require(tidyverse)
require(magrittr)
```


# Summary

This notebook contains a log of all work to create the final pedigree for the 2022 North Santiam Chinook salmon genetic pedigree study.

For each offspring year, assign parentage using two programs: _Cervus_ and _Colony_. Then create a consensus pedigree between the two programs. Finally, combine offspring years into a final pedigree used for downstream analysis.

This is an R notebook. The .html version of this file is a fully rendered and interactive log. To view it, save the html and open in a browse. The .rmd version can be opened within R studio. To reproduce results or edit the analysis: clone the full repository onto tyour local machine and open the r project in rstudio. This will provide all needed data and objects. 



# Data Summary

Input data is available in the directory "parentage_data"  

- This directory contains inputs and outputs for/from Cervus and Colony  
- Subdirectories are organized by offspring year, each subdirectory contains:  
    - all data required to run Cervus and Colony for a given offspring year  
    - unprocessed outputs from Cervus and Colony for a given offspring year  
    - processed Cervus and Colony pedigrees for a given offspring year  
    - note that the same parents occur can occur across multiple offspring years, so all information for a given parent may be spread across multiple subdirectories  
- Input parameters (colony.DAT files, cervus .crv files, logs) reflect the actual runs.    


# Prepare Input Data

## Rationale

Previous files for 2016 - 2019 are ready to use for _Cervus_ and _Colony_, but there is no unified sample naming scheme across years. For example the same individual is named "NS-OP_12_0011" and "NS_12_OP_HOR_M_0011" in the input files for 2017 and 2016 respectively. 

This will wreak havoc with downstream analyses, because there will be no way to readily unify results across offspring years, a step critical for calculation of any statistic calculated for parents, such as total lifetime fitness. 

In this section of the notebook we will change the input files to use a single naming scheme. 

## Log

Changing input file names _in situ_ was going to be very time consuming because nearly each year (and often sex within a year) used a different naming scheme for the same individuals. Instead, I concatenated all of the "final" genotype files into a single dataset, confirmed samples sizes per group using table 3 from the draft report and then reconstructed datasets manually (in excel). I also took the opportunity to unify metadata into a single document. Because each individual only appears once in this dataset, subsetting it to create the inputs for cervus and colony solved the naming convention problem described above.

The document is available at "project_repo/parentage/parentage_data/full_dataset.xlsx"

## Standardize Metadata Format

ODFW/USACE metadata is not standardized. As a consequence, the values used can vary from year to year (e.g. sex is indicated as M, F, male, and female). In the code chunks below, we standardize this data. 

__check data__
First, let's check that we correctly imported this dataset and compare sample sizes against the draft report.
```{r standardize_metadata, message=FALSE, warning=FALSE}
full_dataset <- readxl::read_xlsx("parentage_data/full_dataset.xlsx", sheet = 1, col_types = "text")

tbl_summary(select(full_dataset, type, year), by = type) %>%
  modify_header(label ~ "") %>%
  modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5","stat_6","stat_7") ~ "**type**") %>%
  as_kable_extra() %>%
  kable_classic(full_width = F, html_font = "Arial")# %>%
```

__All looks good here.__ There are only two differences between this table and the draft report:  
(1) New data in 2017 and 2020 added here.  
(2) In 2015 table 3 from the draft report contained an error. Two individuals with no sex information were included in the table, but not considered as possible information in the previous analysis (or this one). I think they got in here by quickly summing up one of Bohn's concatenated genotype table by category, where they were not filtered out to reflect what was done in the analysis.  

Now, let's start standardizing these values.

```{r get_values_to_fix, eval = FALSE}
 full_dataset %>%
  count(geno_sex)

 full_dataset %>%
  count(date)
 
full_dataset %>%
  filter(type == "outplant") %>%
  count(location)
#not too bad to fix, but what about those NAs? do we really not have a release location for some years?
full_dataset %>%
  filter(type == "outplant") %>%
  filter(is.na(location)) %>%
  count(type, year)
# no, we do know, for 149 of 159 missing values, we'll repair these in the data manually, then run ths again. the remaining 10 are truly unknown.
 
full_dataset %>%
  count(origin)

# note here that there are two individuals with no sex info in the new dataset, I removed them to be consistent with previous work


```

```{r, message=FALSE, warning=FALSE}
# now let's fix all but date
full_dataset %<>%
  mutate(origin = case_when(origin %in% c("N", "NOR", "Unclipped") ~ "NOR",
                            origin %in% c("Clipped", "HOR", "Y") ~ "HOR",
                            TRUE ~ NA_character_))

# checked here

full_dataset %<>%
  mutate(location = str_to_lower(location)) %>%
  mutate(location = case_when(location == "breitenbush" ~ "breitenbush river",
                              location == "mainstem/breitenbush" ~ "breitenbush river",
                              TRUE ~ location))
# checked here

full_dataset %<>%
  mutate(geno_sex = case_when(geno_sex == "Female" ~ "F",
                              geno_sex == "Male" ~ "M", 
                              TRUE ~ geno_sex))
  
```

Now to tackle date...
There are a couple issues, the first is that excel saves dates as integers and I didn't tell readxl to read these in as dates. Let's solve that by creating a new dataset to work with.

```{r, message=FALSE, warning=FALSE}
dates <- read_tsv("parentage_data/dates.txt")
dates %<>%
  mutate(date = lubridate::mdy(date))

full_dataset %<>%
  select(-date) %>%
  left_join(dates, by = c("sample_id" = "sample_id")) %>%
  relocate(date)

```

## Above/Below

Whether a sample was outplanted/collected above or below Detroit dam is a piece of imformation we will use throughout assignment, but to keep nomenclature consistent with previous drafts, the "type" variable isn't always clear on this. For example reintro usually means above Minto (below Detroit), but in some years there are reintroductions above Detroit. Also carcass includes both carcasses collected above Detroit and below.

Here I try to change the "type" variable into something that can both be used to filter easily by above and below, but retains the structure from the draft report. Then I create a new variable to describe whether a fish is released above Detroit, or below Big Cliff.

```{r}
#all outplants are above
# all "reintro" is below
# reintro above: aboce
# reintro below: below
# carcass: the fun one, checked all location names against map, all are below except the carcasses labeled as carcass_above, confirmed by Bohn labeled "carcass below" for nearly all sites. an exception is 2017, but these are confirmed below in other metadata

# so we can just label the eceptions here and clean things up: reintros are always below except reintro_above, and carcasses are always below excecpt carcass_above

full_dataset %<>%
  mutate(type = case_when(type == "reintro_below" ~ "reintro",
                          type == "carcass_below" ~ "carcass",
                          TRUE ~ type))
#let's add an above/below variable using this info now
full_dataset %<>%
  mutate(above_below = case_when(type %in% c("reintro", "carcass") ~ "below",
                                 type %in% c("reintro_above", "carcass_above", "outplant") ~ "above"))

```



Excellent, this looks good and we're ready to make some assignments.

# Cervus Log

Actual Cervus run parameters are available for each run using the .crv file in the respective offspring year sub-directory. In general, the following approah was used:  

(1) Estimate allele frequencies using empirical data for each run (offspring year)  
(2) Simulate parentage using empirical allele frequencies and empirical dam and sire sample sizes, 1000 offspring simulated. Additionally:  
    (a) 95% proportion parents sampled  
    (b) 100% loci typed  
    (c) 1% loci mistyped  
    (d) Minimum 7 loci for comparison  
    (e) LOD confidence  
    (f) 80/95 confidence levels  

In some years the number of candidate parents was too large to run cervus (see note 5 in [Assignment Criteria Summary] below). To reduce the number of candidate parents in the final cervus run, first eliminated parents one sex at a time with greater than one mismatch with all offspring. Then used this filtered set of candidate parents to conduct a normal parentage analysis. This should allow LOD scores to be comparable within an offspring year. Some notes on this:  

(1) Used the full dataset allele frequency file for all runs.  
(2) Conducted a new simulation file for each run (cutoffs depend on sample sizes) 
(3) Each analysis in it's own subdirectory. The final datset used for analysis is in subdirectory "filtered_parents"

# Colony Log

Actual Cervus run parameters are available for each run using the .dat or "ProjectInformation.txt" files in the respective offspring year sub-directory. In general, the following approah was used:  

>MATING SYSTEM: MALE/FEMALE POLYGAMY  
MATING SYSTEM II: WITHOUT INBREEDING [default]  
SPECIES: DIOECIOUS, DIPLOID [default]  
LENGTH OF RUN: MEDIUM  
ANALYSIS METHOD: FL-PLS COMBINED (FPLS)  
LIKELIHOOD PRECISION: MEDIUM [default]  
RUN SPECIFICATIONS: NO UPDATE ALLELE FREQUENCY, NO SIBSHIP SCALING, 1 RUN, 1234 RANDOM NUMBER SEED  
SIBSHIP PRIOR: NO PRIOR  
NUMBER OF LOCI: 12  
MARKER TYPES AND ERROR RATES: UPLOAD FILE "MarkerTypeErrorRate-Chinook North Santiam usats"  
ALLELE FREQUENCY: UNKNOWN  
Chance of Sampling a Parent: 95%  
Known Sibs, Excluded Parents etc: 0  

# Assignment Criteria Summary

__Approach__ 

1.	Criteria below are always applied in the order presented here.  
2.	Run colony and cervus again using same version across all offspring years. 
3.	Colony pedigree: 
    a.	Mismatch Filtering (see note 2 though):  
        i.	Remove any parent with >1 mismatch  
        ii.	Remove any trios with >2 mismatches  
    b. remove any impossible trios  
        i.	Both parents must have returned in the same year    
        ii.	Both parents must have been released either above Minto (below Detroit) or above Detroit   
        iii.	If one parent of a trio was collected as a carcass sample, recovery date must be after the release date of the other parent  
4.	Cervus pedigree:  
    a.	Use same parameters as before  
    b.	Combine this dataset with metadata, then use to exclude impossible trios (see 4b above)
    c.	Trios (offspring and two parents): for each offspring choose best trio
        i.	If there is at least one trio with (a) =<2 total mismatches, (b) each parent has =<1 mismatch, (c) significant LOD, then we will make a trio match for this offspring. If there is no trio that meets these criteria, then move on to attempt duo match below.
        ii.	Best trio is chosen by total number of mismatches, ties are broken by LOD score.  
    d.	Duos (offspring and single parent):  
        i.	If no trio available for offspring, we attempt duo match.  
        ii.	If there is a dam or sire with =<1 and significant LOD, we will make a duo assignment.   
        iii.	Best duo is chosen by number of mismatches, ties are broken by LOD score.  
5.	Consensus Pedigree – here we combine the colony and cervus results into the final pedigree used for later analysis  
    a.	Disagreements are defined as any parent that does not agree between the two programs. This includes an assignment in one program and no assignment in the other.  
    c.	If no disagreement, this is the final consensus assignment  
    d.	if either program assigns a trio and disagree:  
        i.	if both assign trio: choose best using total number of mismatches, colony assignment takes precedence in case of tie  
        ii.	if only one trio assigned: this assignment takes precedence  
    e.	if both programs assign duo and disagree:
        i.	parent assignment with fewer mismatches takes precedence, ties go to colony  
    f.	if only one program makes assignment:  
        i.	keep assignment (assignment takes precedence to no assignment)

# Assignment

We make assignments in this section


## Cervus Assignment  

In this section we import data from cervus, clean it up, and apply the assignment criteria to choose the best parentage.

```{r, message=FALSE, warning=FALSE}
#import raw cervus output
cervus_2014 <- read_csv("parentage_data/2014/Cervus/2014_summary.csv")
cervus_2015 <- read_csv("parentage_data/2015/Cervus/2015_out.csv")
cervus_2016 <- read_csv("parentage_data/2016/Cervus/2016_cervus_out.csv")
cervus_2017 <- read_csv("parentage_data/2017/Cervus/filtered_parents/2017_filtered_parents_out.csv")
cervus_2018 <- read_csv("parentage_data/2018/Cervus/2018_out.csv")
cervus_2019 <- read_csv("parentage_data/2019/Cervus/filtered_parents/2019_filtered_out.csv")
cervus_2020 <- read_csv("parentage_data/2020/Cervus/filtered_parents/2020_filtered_out.csv")

# combine years
cervus_raw <- bind_rows(cervus_2014, cervus_2015, cervus_2016, cervus_2017, cervus_2018, cervus_2019, cervus_2020)

# clean up
# cervus does not output machine-parseable, unique column names, instead it uses the index of the column to assign duplicate column names to either mother father or offspring, let's fix this, we'll also take the opportunity to unite names with the colony dataset 

cervus_raw %<>%
  rename(offspring_sample_id = "Offspring ID", 
         offspring_loci_typed ="Loci typed", 
         mother_sample_id = "Candidate mother ID", 
         mother_loci_typed = "Loci typed_1",
         mother_n_comp = "Pair loci compared",
         mother_mm = "Pair loci mismatching",
         mother_lod = "Pair LOD score",
         mother_confidence = "Pair confidence",
         father_sample_id = "Candidate father ID",
         father_loci_typed = "Loci typed_2",
         father_n_comp = "Pair loci compared_1",
         father_mm = "Pair loci mismatching_1",
         father_lod = "Pair LOD score_1" ,
         father_confidence = "Pair confidence_1" ,
         trio_n_comp = "Trio loci compared" ,
         trio_mm = "Trio loci mismatching",
         trio_lod = "Trio LOD score",
         trio_confidence = "Trio confidence") %>%
  select(-c("Pair top LOD","Pair top LOD_1" , "Trio top LOD"  ) ) # remove unused columns
```

```{r cervus meta_data}
# combine with metadata

#offspring
cervus_full <- full_dataset %>%
  select(sample_id, year, type, date) %>%
  rename_with(.fn = ~ paste0("offspring_", .x)) %>%
  right_join(cervus_raw, by = c("offspring_sample_id" = "offspring_sample_id"))

#father
cervus_full <- full_dataset %>%
  select(sample_id, year, type, date, above_below) %>%
  rename_with(.fn = ~ paste0("father_", .x)) %>%
  right_join(cervus_full, by = c("father_sample_id" = "father_sample_id"))

#mother
cervus_full <- full_dataset %>%
  select(sample_id, year, type, date, above_below) %>%
  rename_with(.fn = ~ paste0("mother_", .x)) %>%
  right_join(cervus_full, by = c("mother_sample_id" = "mother_sample_id"))

rm(cervus_raw)

```

```{r}

# trios
cervus_trios <- cervus_full %>%
  filter(mother_mm <= 1 & father_mm <= 1 & trio_mm <= 2 & trio_confidence == "*") %>%    # get rows that meet good trio criteria
  filter(father_year == mother_year) %>% # remove impossibles 
  filter(father_above_below == mother_above_below) %>% # remove impossibles 
  filter(! (str_starts(father_type, "carcass") & !(str_starts(mother_type, "carcass")) & father_date < mother_date)) %>% # remove impossibles 
  filter(! (str_starts(mother_type, "carcass") & !(str_starts(father_type, "carcass")) & father_date > mother_date)) %>% # remove impossibles 
  group_by(offspring_sample_id) %>% # get best row per offspring
  slice_min(trio_mm) %>% # get best row per offspring
  slice_max(trio_lod) # break ties by trio LOD
  
# duos
# for duos the data structure is not "tidy:" we are choosing one parent at a time, yet each row does not represent one parent, but two. So to make things much easier, we will make the dataset a little longer (if not fully long), here each duo in the cervus output is made into its own row

cervus_moms <- cervus_full %>%
  select(-ends_with(c("date", "year", "type", "below"))) %>% #get rid of all the info used to filter impossible trios, don't need it anymore
  select(offspring_sample_id, starts_with("mother")) %>% # now just get the mom info
  rename_with(.fn = ~ str_remove(.x, "mother_"))

cervus_dads <- cervus_full %>%
  select(-ends_with(c("date", "year", "type", "below"))) %>% #get rid of all the info used to filter impossible trios, don't need it anymore
  select(offspring_sample_id, starts_with("father")) %>% # now just get the mom info
  rename_with(.fn = ~ str_remove(.x, "father_"))

cervus_duos <- bind_rows(cervus_dads, cervus_moms, .id = "parent_sex")
rm(cervus_dads, cervus_moms)

cervus_duos %<>%
  filter(!(offspring_sample_id %in% cervus_trios$offspring_sample_id)) %>% # remove offspring that already have assigned trios
  distinct(.keep_all = TRUE) %>% # remove duplicated rows (same parent can appear across multiple rows in cervus output)
  filter(mm <= 1 & confidence == "*") %>% # get duos that match criteria
  group_by(offspring_sample_id) %>% # get the best for each offspring
  slice_min(mm) %>%
  slice_max(lod, with_ties = FALSE) %>%
  select(offspring_sample_id, sample_id, parent_sex) %>% # now reassemble into the previous format
  pivot_wider(id_cols = offspring_sample_id, names_from = parent_sex, values_from = sample_id) %>%
  rename(mother_sample_id="2", father_sample_id = "1") 

cervus_duos %<>%
  left_join(distinct(select(cervus_full, starts_with("father"), offspring_sample_id)), by = c( "father_sample_id" = "father_sample_id" , "offspring_sample_id" = "offspring_sample_id")) %>%
  left_join(distinct(select(cervus_full, starts_with("mother"), offspring_sample_id)), by = c( "mother_sample_id" = "mother_sample_id" , "offspring_sample_id" = "offspring_sample_id")) %>%
  left_join(distinct(select(cervus_full, starts_with("offspring"))), by = c("offspring_sample_id" = "offspring_sample_id"))
  
### now combine into final cervus pedigree
# let's get the info from unassigned offspring
cervus_unassigned_offspring <- cervus_full %>%
  filter(!(offspring_sample_id %in% cervus_trios$offspring_sample_id), !(offspring_sample_id %in% cervus_duos$offspring_sample_id)) %>%
  select(starts_with("offspring")) %>%
  distinct()

cervus_final <- bind_rows(cervus_trios, cervus_duos, cervus_unassigned_offspring) %>%
  relocate(offspring_sample_id, mother_sample_id, father_sample_id, starts_with("mother"), starts_with("father"), starts_with("trio"))


# for the test year (2016) there were offspring with more than one qualifying trios, so we can't be sure this code is fully debugged until we add years, alternatively this may be an attribute of how trio LOD is calculated in cervus
```

## Colony Assignment  

In this section we import data from colony, clean it up and filter out impossible parents.

```{r import, message=FALSE, warning=FALSE}
#import

# colony exports data padded into pseudocolumns using different number of spaces, requires a smart importer, like read_table from readr
colony_2014 <- read_table("parentage_data/2014/Colony/results/2014.BestConfig")
colony_2015 <- read_table("parentage_data/2015/Colony/results/2015.BestConfig")
colony_2016 <- read_table("parentage_data/2016/Colony/results/2016.BestConfig")
colony_2017 <- read_table("parentage_data/2017/Colony/results/2017.BestConfig")
colony_2018 <- read_table("parentage_data/2018/Colony/results/2018.BestConfig")
colony_2019 <- read_table("parentage_data/2019/Colony/results/2019.BestConfig")
colony_2020 <- read_table("parentage_data/2020/Colony/results/2020.BestConfig")


#combine into single dataset
full_colony_raw <- bind_rows(colony_2014, colony_2015, colony_2016, colony_2017, colony_2018, colony_2019, colony_2020)


## lets add the metadata we'll need to filter later (year, type, date) for offspring and each parent
#offspring
full_colony <- full_dataset %>%
  select(sample_id, year, type, date) %>%
  rename_with(.fn = ~ paste0("offspring_", .x)) %>%
  right_join(full_colony_raw, by = c("offspring_sample_id" = "OffspringID"))

#father
full_colony <- full_dataset %>%
  select(sample_id, year, type, date, above_below) %>%
  rename_with(.fn = ~ paste0("father_", .x)) %>%
  right_join(full_colony, by = c("father_sample_id" = "FatherID"))

#mother
full_colony <- full_dataset %>%
  select(sample_id, year, type, date, above_below) %>%
  rename_with(.fn = ~ paste0("mother_", .x)) %>%
  right_join(full_colony, by = c("mother_sample_id" = "MotherID"))

rm(full_colony_raw)
```

```{r colony trios filter}
## trios
# filter trios for impossible parent pairs
colony_trios <- full_colony %>%
  filter(father_year == mother_year) %>%
  filter(father_above_below == mother_above_below) %>%
  filter(! (str_starts(father_type, "carcass") & !(str_starts(mother_type, "carcass")) & father_date < mother_date)) %>%
  filter(! (str_starts(mother_type, "carcass") & !(str_starts(father_type, "carcass")) & father_date > mother_date))

# filter trios for mismatches
colony_trios %<>%
  left_join(select(cervus_full, offspring_sample_id, mother_sample_id, father_sample_id, ends_with("mm")), by = c("offspring_sample_id" = "offspring_sample_id", "mother_sample_id" = "mother_sample_id", "father_sample_id" = "father_sample_id" )) %>%
  filter(father_mm <= 1, mother_mm <=1, trio_mm <=2 ) 
```

```{r colony duos, cache = TRUE}
## duos
# get duos from output
colony_duos <- full_colony %>%
  filter(str_starts(mother_sample_id, "#") | str_starts( father_sample_id, "\\*")) %>%
  filter(!(str_starts(mother_sample_id, "#") & str_starts( father_sample_id, "\\*"))) 
# filter duos for mismatches
# for duos the situation is much more challenging because the singleparents are often not in the cervus output (see note 2 from assignment criteria summary) and we need to count mismatches ourselves

#get simplified genotype data
long_genos <- full_dataset %>%
  select(sample_id, starts_with(c("Ot", "Ssa", "Ogo"))) %>%
  pivot_longer(-sample_id, names_to = "marker", names_pattern = "(.+)\\.\\.\\.")


# debugging example data, do not use in final analysis
#long_genos_sample <- long_genos %>%
#  filter(sample_id %in% c("NS-OP_11_0014", "NS-OP_11_0015", "NS-OP_11_0016", "NS-OP_11_0018", "NS-OP_11_0020", "NS-OP_11_0024", "NS-OP_11_0025", "NS-OP_11_0026", "OtsAC16NSNT_1002", "NS-OP_12_0014"))
#comps_to_make <- data.frame(parent_id = c( "NS-OP_12_0014","NS-OP_11_0015"), offspring_sample_id = c("OtsAC16NSNT_1002", "NS-OP_11_0026"))
#mm_info <- matrix(ncol=3, nrow=nrow(comps_to_make))
#colnames(mm_info) <- c("parent_id","offspring_sample_id", "mm" )

markers <- pull(distinct(long_genos, marker))

# actual data
mm_info <- colony_duos %>%
  mutate(mother_sample_id = case_when(str_starts(mother_sample_id, "#") ~ NA_character_,
                                      TRUE ~ mother_sample_id),
         father_sample_id = case_when(str_starts(father_sample_id, "\\*") ~ NA_character_,
                                      TRUE ~ father_sample_id) ) %>%# turn unknown parents into NAs
  select(offspring_sample_id, father_sample_id, mother_sample_id) %>%
  mutate(parent_id = coalesce(father_sample_id, mother_sample_id)) %>% #combine parents into a single column
  add_column(mm = NA) %>%
  relocate(parent_id, offspring_sample_id, mm)  # relocate columns to format expected by mismatch calculator function



for (i in (1:nrow(mm_info))){
  match_marker = 0
  for (m in markers){
    match_marker = match_marker + (sum(pull(long_genos[long_genos$marker == m & long_genos$sample_id == as.character(mm_info[i,1]), 3] ) %in% pull(long_genos[long_genos$marker == m & long_genos$sample_id == as.character(mm_info[i,2]), 3] )) >=1)
    #print(match_marker)
  }
  mismatch = 12-(match_marker)
  #mm_info[i,1] <- as.character(comps_to_make[i,1]) #print parent to results matrix
  #mm_info[i,2] <- as.character(comps_to_make[i,2]) #print offspring to results matrix
  mm_info[i,3] <- mismatch# print number of mismatches
}

# now that we have mismatches, remove any duos with >1 mismatch
colony_duos <- mm_info %>%
  filter(mm <=1) %>%
  select(offspring_sample_id, father_sample_id, mother_sample_id)


## unassigned
# collect unassigned info
colony_unassign <- full_colony %>%
  select(offspring_sample_id) %>%
  distinct(offspring_sample_id) %>%
  filter(!(offspring_sample_id %in% c(colony_duos$offspring_sample_id, colony_trios$offspring_sample_id)))


# combine filtered trios with duos
colony_final <- bind_rows( select(colony_trios, offspring_sample_id, father_sample_id, mother_sample_id), colony_duos, colony_unassign)

# here we should consider if we want to include NA rows for offspring with unassigned parents, will depend on our approach to consensus pedigree code, which is currently unwritten, to keep all just remove the last line of the duos code section
```

```{r, eval = FALSE}
# let's check that the final colony pedigree matches expectation

full_colony %>%
  filter(! (offspring_sample_id %in% colony_final$offspring_sample_id)) %>%
  View(.)

#only cheked for 2016, but it worked perfectly in that year, come back and check again with the full dataset before finalizing assignments. 
```

## Consensus Assignment 


```{r}
## collect and simplify mm info for tie breaking
#colony first
colony_for_consensus <- colony_final %>%
  rename(colony_mother= mother_sample_id, colony_father = father_sample_id) %>%
  left_join(select(all_of(colony_trios), offspring_sample_id, colony_mother= mother_sample_id, colony_father = father_sample_id, colony_trio_mm = trio_mm), by = c("offspring_sample_id" = "offspring_sample_id", "colony_mother" = "colony_mother", "colony_father" = "colony_father")) %>%
  left_join(select(all_of(mm_info), offspring_sample_id, colony_mother= mother_sample_id, colony_father = father_sample_id, colony_father_mm = mm, colony_mother_mm = mm), by = c("offspring_sample_id" = "offspring_sample_id", "colony_mother" = "colony_mother", "colony_father" = "colony_father")) %>%
  mutate(colony_father_mm = case_when(is.na(colony_father) ~ NA_real_,
                               TRUE ~ colony_father_mm),
         colony_mother_mm = case_when(is.na(colony_mother) ~ NA_real_,
                               TRUE ~ colony_mother_mm)) 

#now cervus
cervus_for_consensus <- cervus_final %>%
  select(offspring_sample_id,  cervus_mother = mother_sample_id, cervus_mother_mm = mother_mm, cervus_father = father_sample_id, cervus_father_mm = father_mm, cervus_trio_mm = trio_mm)

#now combine
consensus_df <- colony_for_consensus %>%
  left_join(cervus_for_consensus, by = c("offspring_sample_id" = "offspring_sample_id"))

```

```{r}
# now we apply the consensus rules

# let's get trio information to make some of this easier/faster
consensus_df %<>%
  replace_na(list(colony_mother = "none", colony_father = "none", cervus_father = "none", cervus_mother = "none")) %>%
  mutate(colony_trio = case_when((colony_mother != "none" & colony_father != "none") ~ 1,
                                 TRUE ~ 0),
         cervus_trio = case_when((cervus_mother != "none" & cervus_father != "none") ~ 1,
                                 TRUE ~ 0)
         )

# total agreement

final_pedigree <- consensus_df %>%
  mutate(final_mother = case_when((colony_mother == cervus_mother & colony_father == cervus_father) ~ colony_mother, 
                                  (colony_trio == 1 & cervus_trio == 1) & (cervus_trio_mm < colony_trio_mm) ~ cervus_mother,
                                  (colony_trio == 1 & cervus_trio == 1) & (cervus_trio_mm > colony_trio_mm) ~ colony_mother,
                                  (colony_trio == 1 & cervus_trio == 1) ~ colony_mother,
                                  (colony_trio == 1 & cervus_trio == 0) ~ colony_mother,
                                  (colony_trio == 0 & cervus_trio == 1) ~ cervus_mother,
                                  (colony_mother == "none" & cervus_mother != "none" ) ~ cervus_mother, 
                                  (colony_mother != "none" & cervus_mother == "none") ~ colony_mother,
                                  (colony_mother == "none" & cervus_mother == "none") ~ "none",
                                  (colony_mother_mm <= cervus_mother_mm) ~ colony_mother,
                                  (colony_mother_mm > cervus_mother_mm) ~ cervus_mother)) %>%
  mutate(final_father = case_when((colony_father == cervus_father & colony_father == cervus_father) ~ colony_father, 
                                  (colony_trio == 1 & cervus_trio == 1) & (cervus_trio_mm < colony_trio_mm) ~ cervus_father,
                                  (colony_trio == 1 & cervus_trio == 1) & (cervus_trio_mm > colony_trio_mm) ~ colony_father,
                                  (colony_trio == 1 & cervus_trio == 1) ~ colony_father,
                                  (colony_trio == 1 & cervus_trio == 0) ~ colony_father,
                                  (colony_trio == 0 & cervus_trio == 1) ~ cervus_father,
                                  (colony_father == "none" & cervus_father != "none" ) ~ cervus_father, 
                                  (colony_father != "none" & cervus_father == "none") ~ colony_father,
                                  (colony_father == "none" & cervus_father == "none") ~ "none",
                                  (colony_father_mm <= cervus_father_mm) ~ colony_father,
                                  (colony_father_mm > cervus_father_mm) ~ cervus_father)) %>%
  relocate(final_mother, final_father, colony_mother, colony_father, cervus_mother, cervus_father)

```


```{r, eval = FALSE}
# it looks like the code above wasn't perfect, there's a case where each program assigns a single good parent, but of opposite sex. Reading the description of the assignment criteria, one would think the final retained parentage would be a single parent, but the code above will produce a trio. This is not good becaue it can also produce impossible parentages. This only happens in 0.07% of the assignments, so it was easy to miss in the first round of debugging.

# this code chunk does some testing/debugging of the fix, don't run in the final

# cases of this are below

 final_pedigree %>%
  filter(final_father != "none", final_mother != "none", cervus_father == "none", cervus_mother != "none", colony_father != "none", colony_mother == "none") %>% View(.)

 final_pedigree %>%
  filter(final_father != "none", final_mother != "none", cervus_father != "none", cervus_mother == "none", colony_father == "none", colony_mother != "none") %>% View(.)

#fortunately this is fixable. We can use the same dataframe from above and simply choose a single parent on the basis of MM, with ties going to colony. 


fp2 <- final_pedigree %>%
  mutate(final_father = case_when((final_father != "none" & final_mother != "none" & cervus_father == "none" & cervus_mother != "none" & colony_father != "none" & colony_mother == "none" & cervus_mother_mm < colony_father_mm) ~ "none",
                                  TRUE ~ final_father)) %>%
  mutate(final_mother = case_when((final_father != "none" & final_mother != "none" & cervus_father == "none" & cervus_mother != "none" & colony_father != "none" & colony_mother == "none" & cervus_mother_mm >= colony_father_mm) ~ "none",
                                  TRUE ~ final_mother)) %>%
  mutate(final_mother = case_when((final_father != "none" & final_mother != "none" & cervus_father != "none" & cervus_mother == "none" & colony_father == "none" & colony_mother != "none" & (colony_mother_mm > cervus_father_mm)) ~ "none",
                                  TRUE ~ final_mother)) %>%
  mutate(final_father = case_when((final_father != "none" & final_mother != "none" & cervus_father != "none" & cervus_mother == "none" & colony_father == "none" & colony_mother != "none" & (colony_mother_mm <= cervus_father_mm)) ~ "none",
                                  TRUE ~ final_father))



# check if this worked
anti_join(fp2, final_pedigree) %>% View(.) 
# yes, for the fathers, we should have only changed on eof the three cases

```

```{r}
# now we implement the fix from above
final_pedigree %<>%
  mutate(final_father = case_when((final_father != "none" & final_mother != "none" & cervus_father == "none" & cervus_mother != "none" & colony_father != "none" & colony_mother == "none" & cervus_mother_mm < colony_father_mm) ~ "none",
                                  TRUE ~ final_father)) %>%
  mutate(final_mother = case_when((final_father != "none" & final_mother != "none" & cervus_father == "none" & cervus_mother != "none" & colony_father != "none" & colony_mother == "none" & cervus_mother_mm >= colony_father_mm) ~ "none",
                                  TRUE ~ final_mother)) %>%
  mutate(final_mother = case_when((final_father != "none" & final_mother != "none" & cervus_father != "none" & cervus_mother == "none" & colony_father == "none" & colony_mother != "none" & (colony_mother_mm > cervus_father_mm)) ~ "none",
                                  TRUE ~ final_mother)) %>%
  mutate(final_father = case_when((final_father != "none" & final_mother != "none" & cervus_father != "none" & cervus_mother == "none" & colony_father == "none" & colony_mother != "none" & (colony_mother_mm <= cervus_father_mm)) ~ "none",
                                  TRUE ~ final_father))

```


```{r, eval = FALSE}
# quick stats 
final_pedigree %>%
  summarise(unassigned_n = sum(final_mother == "none" & final_father == "none", na.rm = TRUE))

#write_tsv(final_pedigree, "final_pedigree.txt")
```


## Save Pedigree

Let's save these results as txt files and r objects that can be used in subsequent analyses.

Metadata first. We've made some changes to the metadata, let's save a final version of this information
```{r, eval = FALSE}

#everything 
save(full_dataset, file= "parentage_results/full_dataset.R")
write_tsv(full_dataset, "parentage_results/full_dataset.txt")

# metadata only
meta_data <- full_dataset %>%
  select(-starts_with(c("Ot", "Ogo", "Ssa")))
save(meta_data, file= "parentage_results/meta_data.R")
write_tsv(meta_data, "parentage_results/meta_data.txt")  

```

Now the pedigree.
```{r, eval = FALSE}
pedigree <- final_pedigree %>%
  select(offspring_sample_id, mother = final_mother, father = final_father)

save(pedigree, file = "parentage_results/pedigree.R")
write_tsv(pedigree, "parentage_results/pedigree.txt")

```

